\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for Vietnamese and International Students

### 2023-06-19 phd-cs-us.tex
updates

\begin{commentbox}[Vu]
One of the reasons I create this document is that my (non-Vietnamese) colleagues at GMU are interested in 
recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (e.g., each year our CS program receives more than 350 PhD applications, most of which are international but only 3--4 are from Vietnam). In general the number of
PhD applications from Vietnam to US universities is few and  more would be very welcomed. 
\end{commentbox}

@@

\begin{commentbox}[Hakan]
At GMU, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline, and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as "optional", it is mandatory and they should plan for submitting all those.  
\end{commentbox}
\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

While you might not have control over LORs \hyperref[sec:your-school]{where your go to school}, you do over your
statement! So write it well because we take it seriously.
It is something that can make your application unique and stand out, which can make you "fit"  the CS program you're applying to, even if you don't have very strong research experience.
A well-written LOR also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
%is important because if you need (GTA) funding, it will provide evidence
%that you can teach and communicate with students.

@@

\begin{commentbox}[Vu]
TA and RA at GMU have similar benefits in tuition waiving and insurance.  For stipend, the college and department will set a 9-month graduate assistant stipend.  TA and RA will usually be that amount (TA will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof.). 
\tcblower
Having health insurance is required at many US universities.  So do not think that you're young and healthy and ignore insurance.  At GMU, and at most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

\subsection{Tenured or tenure-track faculty? Who do you choose?}
\begin{commentbox}[Thanh]
In my opinion, having a well-suited advisor is crucial for a successful PhD and research career. One effective approach to finding a suitable professor is by working with a professor during your undergraduate studies. An exemplary instance is VinAI's residency program, where residents collaborate with professors from the US for two years before applying to PhD programs. Many VinAI residents have achieved remarkable results and gained admission to prestigious US universities. Unfortunately, VinAI's resident program is limited to AI research. In other fields, e.g. SE,
Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants?
\end{commentbox}
\begin{wrapfigure}{r}{0.48\textwidth}
    \vspace{-0.3in}
      \begin{center}
        \includegraphics[width=0.48\textwidth]{c8.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
The short answer is that tenured-track faculty (e.g., assistant professors) are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding as tenured faculty.

@@

@@

\subsection{Should you contact a US professor? How to do so "correctly"?}

\tvn{TODO: How to write an email that will get a reply?}

@@

\subsection{Is an MS degree required for admission to PhD?}
No. In fact, student with BS can get MS degree "along the way" to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.

@@

\begin{wrapfigure}{r}{0.28\textwidth}
    \vspace{-0.2in}
      \begin{center}
        \includegraphics[width=0.28\textwidth]{c4a.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
    Typically, 5-7 years for PhD in CS at US universities.  The first 2 years you spend on coursework, finding an adviser, learning research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published.  The last 1-2 years you continue to publish, write and defend your dissertation, and look for job. The PhDComics figure on the right shows the "ambition" level of a PhD student over their years of study (they miss the 6-7th Year where the ambition is "Just get me graduate").

>>>>

\begin{commentbox}[Vu]
One of the reasons I create this document is that my colleagues at GMU are interested in 
recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (e.g., each year our CS program receives more than 350 PhD applications, most of which are international but only 3--4 are from Vietnam). In general the number of
PhD applications from Vietnam to US universities is few and  more would be very welcomed. 
\end{commentbox}

@@

\begin{commentbox}[Hakan]
At GMU, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline, and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as ``optional'', it is mandatory and they should plan for submitting all those.  
\end{commentbox}
\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

While you might not have control over LORs \hyperref[sec:your-school]{where your go to school}, you do over your
statement! So write it well because we take it seriously.
In your statement, you have the opportunity to make  make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written LOR also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
%is important because if you need (GTA) funding, it will provide evidence
%that you can teach and communicate with students.

@@

\begin{commentbox}[Vu]
TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for 9-month graduate assistant stipend.  TA will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof. 
\tcblower
Having health insurance is required at many US universities.  So do not think that you're young and healthy and ignore insurance.  At GMU, and at most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

\subsection{Tenured or tenure-track faculty? Who do you choose?}

\begin{wrapfigure}{r}{0.48\textwidth}
    \vspace{-0.3in}
      \begin{center}
        \includegraphics[width=0.48\textwidth]{c8.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
The short answer is that tenured-track faculty (e.g., assistant professors) are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding as tenured faculty.

@@

\begin{commentbox}[Thanh]
    In my opinion, having a well-suited advisor is crucial for a successful PhD and research career. One effective approach to finding a suitable professor is by working with a professor during your undergraduate studies. An exemplary instance is VinAI's residency program, where residents collaborate with professors from the US for two years before applying to PhD programs. Many VinAI residents have achieved remarkable results and gained admission to prestigious US universities. Unfortunately, VinAI's resident program is limited to AI research. In other fields, e.g. SE,
    Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants?
    \end{commentbox}

@@

\subsection{Should you contact a US professor? How to do so ``correctly''?}

If you want to contact a prof. to \emph{ask about your admission chance}, then \textbf{don't}. We don't know, we can't answer because as explained throughout this document, we don't make individual decisions and might not even be assigned to evaluate your application.  It is the same as sending a paper draft to a journal editor and ask them if your paper will be accepted.

If you want to contact a prof. to ask about \emph{research opportunities}, then \emph{may be}.  First, you should read the prof's website, see if they are saying something about contacting them. Many profs. indicate what you should do to contact them (e.g., specific email subjects). In general cases, you need to \emph{customize your email}. For example, read their papers, and see if you are interested in their work.  If you are, then you can send them an email talking how/why their work would match yours (give details and evidence, don't just say I am a software developer and therefore match with your interests in software design) and ask about research opportunities.

@@

\subsection{Is an MS degree required for admission to PhD?}
No. In fact, student with BS can get MS degree ``along the way'' to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.

If you have an MS then some course work \emph{might be} transferred for course credits, which save a bit of time. But overall don't count on this, especially if your MS is not from the US.

% \begin{commentbox}[Thanh]
% \end{commentbox}

@@

\begin{wrapfigure}{r}{0.28\textwidth}
    \vspace{-0.2in}
      \begin{center}
        \includegraphics[width=0.28\textwidth]{c4a.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
    Typically, 5--7 years for PhD in CS in the US.  This can be longer than CS PhD at universities in other countries, which might require MS first (recall that US PhD programs do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students in the US also often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

    The first 2 years you spend on coursework, finding an adviser, learning research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job. 
    In many cases you might take a summer or two off to do internship to get additional research opportunities.
    The PhDComics figure on the right shows the ``ambition'' level of a PhD student over their years of study (they miss the 6-7th Year where the ambition is \emph{``Just let me graduate''}).
